Understanding Factors Contributing to a Successful Himalayan Expedition

### 2024-04-19 paper/paper.qmd
Trace plots added

#### Workspace setup ####
library(broom.mixed)
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
library(modelsummary)
library(here)
library(kableExtra)
library(arrow)

@@

@fig-counts displays the counts for various variables I will be modelling. We see that there are significantly higher men than women. This shows the existence of a gender imbalance skewed towards men in mountaineering in the Himalayas. Additionally, most expedition members opt not to attempt solo ascents and fall within the middle-aged range. Lastly, we see the peak expedition seasons are the more pleasant autumn and spring over the extreme seasons like summer and winter. These count disparities should be considered when analysing success proportions.

@@

@fig-proportions shows the proportion of expedition members who had a successful outcome or died during their expedition. For sex, we see that more men had successful expeditions compared to women. Moreover, individuals who embark on solo expeditions have a higher mortality rate, potentially explaining the difference in the number of people attempting solo ascent observed earlier. Some other trends that can be observed are that success rates decline with age, and winter contributes to the most deaths. The success and death proportions based on the height of the peak are very varied and do not follow a specific trend. This goes to show that just the height of the peak alone doesn't define the probability of having a successful summit but there are other factors that contribute to it.

@@

@@

# Discussion {#sec-discussion}
## Male dominance in mountaineering
Mountaineering, as evident by the data, presents a significant gender gap, with men outnumbering women by nearly sixfold in expedition participation. @fig-counts shows us the counts of our independent variables and if we look at sex, we notice there are almost six times more male expedition members. This gender gap underscores the sport's male-dominated nature.

@@

Firstly, some success probabilities should be viewed with skepticism due to potential inaccuracies. Discrepancies in the counts of different categories may lead to imprecise success probabilities. For instance, the number of individuals attempting a solo ascent was significantly lower than those in group ascents. However, our model coefficients indicate for solo ascents was higher. This could create a misconception that solo ascents have better chances of success, potentially influencing more people to choose this option without considering other contributing factors. These probabilities may be influenced by the limited sample size of solo ascents, which lacks diverse data to account for confounding variables. There might be other factors which play a role like will power. Individuals opting for solo expeditions likely possess high levels of determination which can be a confounding factor playing into their success.

Additionally, while we focused on focused on a few factors, there are numerous other variables that could impact the likelihood of a successful summit. Nationality and group size were two factors which were available in the raw data and I believed were factors which could have played a significant difference. These were later removed due to the lack of consistency in those variable’s. For instance, the way the number of members were reported differed from record to record. As the dataset also included the people summiting the other side of border peaks, some countries reported this number differently than others. For expeditions in Nepal, this was usually the number of foreigners listed on the expedition permit whereas in China, this was usually the number of non-hired members.

## Future Directions
Moving forward, additional research is necessary to address the limitations discussed and improve the dataset to enhance our understanding of factors contributing to a successful summit. Including more diverse information can help uncover additional factors influencing a successful summit. Some of these factors that can be explored would be nationality of the person, attempt number, a metric for that could measure their level of preparedness and group size.

@@

# Appendix

@@

library(bayesplot)

pp_check(expeditions_model) +
  theme(legend.position = "bottom")

```

@@

library(broom)
library(broom.mixed)
library(knitr)
library(modelsummary)
library(purrr)
library(rstanarm)
library(testthat)
library(tidyverse)

posterior_vs_prior(expeditions_model) +
  theme_minimal() +
  theme(axis.text.x=element_text(hjust=1, vjust=0.5),
        text=element_text(size=8)) +
  guides(color = guide_legend(ncol = 6))+
  theme(legend.position = "bottom") +
  coord_flip()
```
@fig-post and @fig-prior-post are used to understand assess the quality of our model.

@@

## Markov chain Monte Carlo (MCMC) Convergence Check
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace
#| fig-cap: "Trace plot"

layout(matrix(1:6, ncol = 3))
plot(expeditions_model, "trace")

```

@@

plot(expeditions_model, "rhat")
```
My model, modelled using `rstanarm` [@citeRstanarm] uses a sampling algorithm called Markov chain Monte Carlo (MCMC) to obtain samples from the posterior distributions of interest. @fig-trace and @fig-rhat are used to check for the existence of signs that the algorithm ran into issues. @fig-trace does not suggest anything out of the ordinary and everything in @fig-rhat is close to 1 suggesting no problem as well.

>>>>

#### Workspace setup ####
library(broom.mixed)
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
library(here)
library(kableExtra)
library(arrow)
library(rstanarm)

@@

@fig-counts displays the counts for various variables I will be modelling. The plot shows there are significantly higher men than women. This shows the existence of a gender imbalance skewed towards men in mountaineering in the Himalayas. Additionally, most expedition members opt not to attempt solo ascents and fall within the middle-aged range. Lastly, it can be seen that the peak expedition seasons are the more pleasant autumn and spring over the extreme seasons like summer and winter. These count disparities should be considered when analysing success proportions.

@@

@fig-proportions shows the proportion of expedition members who had a successful outcome or died during their expedition. For sex, it is observed that more men had successful expeditions compared to women. Moreover, individuals who embark on solo expeditions have a higher mortality rate, potentially explaining the difference in the number of people attempting solo ascent observed earlier. Some other trends that can be observed are that success rates decline with age, and winter contributes to the most deaths. The success and death proportions based on the height of the peak are very varied and do not follow a specific trend. This goes to show that just the height of the peak alone doesn't define the probability of having a successful summit but there are other factors that contribute to it.

@@

Model coefficients with their standard errors, posterior checks for the model and convergence checks for the model can be found in @sec-appendix

@@

# Discussion {#sec-discussion}
## Male dominance in mountaineering
Mountaineering, as evident by the data, presents a significant gender gap, with men outnumbering women by nearly sixfold in expedition participation. @fig-counts shows the counts of the independent variables and it is observed that there are almost six times more male expedition members. This gender gap underscores the sport's male-dominated nature.

@@

Firstly, some success probabilities should be viewed with skepticism due to potential inaccuracies. Discrepancies in the counts of different categories may lead to imprecise success probabilities. For instance, the number of individuals attempting a solo ascent was significantly lower than those in group ascents. However, my model coefficients indicate for solo ascents was higher. This could create a misconception that solo ascents have better chances of success, potentially influencing more people to choose this option without considering other contributing factors. These probabilities may be influenced by the limited sample size of solo ascents, which lacks diverse data to account for confounding variables. There might be other factors which play a role like will power. Individuals opting for solo expeditions likely possess high levels of determination which can be a confounding factor playing into their success.

Additionally, while I focused on focused on a few factors, there are numerous other variables that could impact the likelihood of a successful summit. Nationality and group size were two factors which were available in the raw data and I believed were factors which could have played a significant difference. These were later removed due to the lack of consistency in those variable’s. For instance, the way the number of members were reported differed from record to record. As the dataset also included the people summiting the other side of border peaks, some countries reported this number differently than others. For expeditions in Nepal, this was usually the number of foreigners listed on the expedition permit whereas in China, this was usually the number of non-hired members.

## Future Directions
Moving forward, additional research is necessary to address the limitations discussed and improve the dataset to enhance the understanding of factors contributing to a successful summit. Including more diverse information can help uncover additional factors influencing a successful summit. Some of these factors that can be explored would be nationality of the person, attempt number, a metric for that could measure their level of preparedness and group size.

@@

# Appendix {#sec-appendix}

@@

pp_check(expeditions_model) +
  theme(legend.position = "bottom")
```

@@

posterior_vs_prior(expeditions_model) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 0.5),
    text = element_text(size = 8)
  ) +
  guides(color = guide_legend(ncol = 6)) +
  theme(legend.position = "bottom") +
  coord_flip()
```
@fig-post and @fig-prior-post are used to understand assess the quality of my model.

@@

## Markov chain Monte Carlo (MCMC) Convergence Check
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace1
#| fig-cap: "Trace plot"
#| fig-subcap: ["Intercept", "sexM", "soloTRUE", "seasonsSpring", "seasonsSummer", "seasonsWinter", "age_rangeUnder 18",  "age_range31-40", "age_range41-50", "age_range51-60", "age_range61-70", "age_range71-80", "age_range81-90"]
#| layout-ncol: 3

plot(expeditions_model, "trace", "(Intercept)")
plot(expeditions_model, "trace", "sexM")
plot(expeditions_model, "trace", "soloTRUE")
plot(expeditions_model, "trace", "seasonsSpring")
plot(expeditions_model, "trace", "seasonsSummer")
plot(expeditions_model, "trace", "seasonsWinter")
plot(expeditions_model, "trace", "age_rangeUnder 18")
plot(expeditions_model, "trace", "age_range31-40")
plot(expeditions_model, "trace", "age_range41-50")
plot(expeditions_model, "trace", "age_range51-60")
plot(expeditions_model, "trace", "age_range61-70")
plot(expeditions_model, "trace", "age_range71-80")
plot(expeditions_model, "trace", "age_range81-90")
```
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace2
#| fig-cap: "Trace plot"
#| fig-subcap: ["height_range5750 - 6099", "height_range6100 - 6449", "height_range6450 - 6799", "height_range6800 - 7149", "height_range7150 - 7499", "height_range7500 - 7849", "height_range7850 - 8199", "height_range8200 - 8549", "height_range8550 - 8900"]
#| layout-ncol: 3

plot(expeditions_model, "trace", "height_range5750 - 6099")
plot(expeditions_model, "trace", "height_range6100 - 6449")
plot(expeditions_model, "trace", "height_range6450 - 6799")
plot(expeditions_model, "trace", "height_range6800 - 7149")
plot(expeditions_model, "trace", "height_range7150 - 7499")
plot(expeditions_model, "trace", "height_range7500 - 7849")
plot(expeditions_model, "trace", "height_range7850 - 8199")
plot(expeditions_model, "trace", "height_range8200 - 8549")
plot(expeditions_model, "trace", "height_range8550 - 8900")
```

@@

plot(expeditions_model, "rhat")
```
My model, modelled using `rstanarm` [@citeRstanarm] uses a sampling algorithm called Markov chain Monte Carlo (MCMC) to obtain samples from the posterior distributions of interest. @fig-trace1, @fig-trace2 and @fig-rhat are used to check for the existence of signs that the algorithm ran into issues. @fig-trace1 and @fig-trace2 does not suggest anything out of the ordinary and everything in @fig-rhat is close to 1 suggesting no problem as well.
